### 2016-08-25 appendix.tex
page 1 to 10 corrections
more corrections

In case the \texttt{losetup -P} is not available given that a version
higher or equal than 2.21 for \texttt{util-linux} is neither not available,
a possible option is to manually set all the loopback devices used during the
whole procedure with \texttt{mknod}. To do this, the alternative commands are:

@@

In case the \texttt{umount --recursive} is not available given that a version
higher or equal than 2.22 for \texttt{util-linux} is neither not available,
a possible option is to manually umount all the partitions used during the
\texttt{chroot} environment. The directories must be unmounted in the reverse
order as when they were mounted. These commands are:

@@

\begin{lstlisting}[]
# pip install --upgrade paramiko
...
Sucesfully installed cffi-1.7.0 cryptography-1.4 paramiko-2.0.1
# pip show paramiko
...
Name: paramiko
Version: 2.0.1
...
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

>>>>

In case the \texttt{losetup -P} command shows the \texttt{invalid option}
message, an alternative is to manually set all the loopback devices
used during the whole procedure with \texttt{mknod}. To do this, the
alternative commands are:

@@

%The recursive command (\texttt{----recursive}) is not valid for \texttt{umount}
%is not valid for \texttt{util-linux} version 2.22
%or later. An alternative option is to manually unmount all the partitions used
%during the \texttt{chroot} environment. The directories must be unmounted in
%the reverse order as when they were mounter. These commands are:

%The --recursive parameter is not part of umount until 
%\texttt{umount ----recursive} is not part of \texttt{util-linux} version 2.22

In case the \texttt{umount ----recursive} command shows the \texttt{invalid option}
message, an alternative is to manually unmount all the partitions used
during the \texttt{chroot} environment. The directories must be unmounted in
the reverse order as when they were mounter. These commands are:
%In case the \texttt{umount ----recursive} command fails given that a version
%higher than 2.22 for \texttt{util-linux} is not available,
%a possible option is to manually umount all the partitions used during the
%\texttt{chroot} environment. The directories must be unmounted in the reverse
%order as when they were mounted. These commands are:

@@

\begin{lstlisting}[]
# pip install --upgrade paramiko
...
Successfully installed cffi-1.7.0 cryptography-1.4 paramiko-2.0.1
# pip show paramiko
...
Name: paramiko
Version: 2.0.1
...
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

### 2016-08-25 conclusions.tex
more corrections

\label{sec:conclusions}
Observing the expectation of the \ac{IoT} and lack for a low-cost, easy to configure testbed that enables reproducible results for mechanisms to address problems in this area, we provide a in-depth description of the new Aalborg University's \ac{Raspi} testbed for network coding evaluation and how to guarantee replicability and scaling managing of this system. The description shows how to set up interconnected \ac{Raspi}s with SD memory cards for local storage, a Raspbian Lite image, network connectivity and proper system administration privileges. We hope this work permits researchers to replicate setups and scenarios for evaluating their strategies in a rapid and manageable way. Future work in the use of \ac{Raspi} devices will focus on expanding the setup and automation of tasks to run the testbed, configure specified network topologies (e.g., with specific connectivity or packet loss ratios), reserve the use of these sub-networks for running tailored experiments and opening the use of the testbed beyond our team at Aalborg University.

>>>>

\label{sec:conclusions}
Observing the expectation of the \ac{IoT} and lack for a low-cost, easy to configure testbed that enables reproducible results for mechanisms to address problems in this area, we provide a in-depth description of the new Aalborg University's \ac{Raspi} testbed for network coding evaluation and how to guarantee replicability and scaling managing of this system. The description shows how to set up interconnected \ac{Raspi}s with memory cards for local storage, a Raspbian Lite image, network connectivity and proper system administration privileges. We hope this work permits researchers to replicate setups and scenarios for evaluating their strategies in a rapid and manageable way. Future work in the use of \ac{Raspi} devices will focus on expanding the setup and automation of tasks to run the testbed, configure specified network topologies (e.g., with specific connectivity or packet loss ratios), reserve the use of these sub-networks for running tailored experiments and opening the use of the testbed beyond our team at Aalborg University.

### 2016-08-25 cross_compilation.tex
more corrections

\label{sec:cross_compilation}
An important case of a computational expensive task is to compile software
packages and large libraries. Given the computing capabilities of the
\ac{Raspi}, such tasks can be challenging if not prohibitive in terms of
\ac{CPU}, memory or space usage and / or compilation time. Thus, in
this Section we present a procedure of how to cross-compile C++
source code from the testbed administrator \ac{PC} for the \ac{ARM}
architecture of the \ac{Raspi}s. By doing this, we take advantage
of the (typically) much higher computing power of the testbed
administrator \ac{PC} in order to save time and computational resources.
Hence, we give an example of compiling a simple C++ program and
copying the generated binaries with \ac{SSH} to run locally
on a \ac{Raspi}.

@@

To use Kodo for research, it is required to obtain a research free license.
To do this, only a request form needs to be filled
in~\cite{steinwurflicenselink} and wait for grant access. Once
the access for Kodo has been granted, the source code can be pulled from
its Git repositories to be compiled. Thus, assuming that testbed
administrator already has access, we clone the \texttt{kodo-cpp}
repository locally in \texttt{\${CODEDIR}} and change directory into
the repository by doing:

>>>>

\label{sec:cross_compilation}
An important case of a computational expensive task is to compile software
packages and large libraries. Given the computing capabilities of the
\ac{Raspi}, such tasks can be challenging if not prohibitive in terms of
\ac{CPU}, memory or space usage and / or compilation time. Thus, in
this section we present a procedure of how to cross-compile C++
source code from the testbed administrator \ac{PC} for the \ac{ARM}
architecture of the \ac{Raspi}s. By doing this, we take advantage
of the (typically) much higher computing power of the testbed
administrator \ac{PC} in order to save time and computational resources.
Hence, we give an example of compiling a simple C++ program and
copying the generated binaries with \ac{SSH} to run locally
on a \ac{Raspi}.

@@

To use Kodo for research, it is required to obtain a research free license.
To do this, only a request form needs to be filled
in~\cite{steinwurflicenselink} and wait for it to be processed by the Kodo
developers. Once
the access for Kodo has been granted, the source code can be pulled from
its Git repositories to be compiled. Thus, assuming that the testbed
administrator already has access, we clone the \texttt{kodo-cpp}
repository locally in \texttt{\${CODEDIR}} and change directory into
the repository by doing:

### 2016-08-25 image_setup.tex
more corrections

\label{sec:image_setup}
In this Section, we review the steps to create a common \ac{OS} image for
all the \ac{Raspi}s. The image setup is composed of three major steps:
Select and download the \ac{OS} image file, alter the image structure and
configure the \ac{OS} files. We proceed to detail all these steps providing
brief discussions to our setup choices when required. To perform these
steps, we indicate with command-line blocks the required sequential
commands to be typed by the testbed administrator in his / her \ac{PC}
to obtain the desired setting. In all the following sections, the number
sign (\#) and dollar sign (\$) will be used in the command blocks in the
paper to indicate if an operation needs to be run with root permissions or
common user permissions, respectively. The signs will be posted as the
terminal prompt in the command blocks.

@@

In this code block, the \texttt{\$\{URL\}} and \texttt{\$\{IMAGE\}}
variables specify where the Linux bundle is located and
\texttt{\$\{WORKDIR\}} specifies a working directory where the Raspbian
Lite bundle will be downloaded and customized. Notice that even though we
use the \$ and \# signs in the shell, in general these signs will be
particular to the testbed administrator \ac{OS} shell. Next, we create
the working directory and go to it with the change directory (\texttt{cd}) command. To
download the image, we utilize the \texttt{wget} command before unpacking
the \texttt{zip} file as follows:

@@

After Raspbian Lite has been unpacked, there should be an \texttt{.img}
file in the working directory \texttt{\$\{WORKDIR\}}. Here, \texttt{fdisk} can
be used to display the content of the image. To do this and obtain
administrative information, we parse the arguments \texttt{-u sectors}
to display the sizes in sectors and \texttt{-l} to display the partitions
within the image. The \texttt{fdisk} command should output to the terminal
something similar to:

@@

\begin{lstlisting}[]

Command (m for help): Partition number (1-4):
Command (m for help): Partition type:
   p   primary (1 primary, 0 extended, 3 free)
   e   extended
Select (default p): Using default response p
Partition number (1-4, default 2): Using default value 2
First sector (2048-4806655, default 2048): Last sector, +sectors or
+size{K,M,G} (137216-4806655, default 4806655): Using default value 4806655

@@

\subsection{Loopback Device Setup}
After successfully resizing the image file, we use a loopback device to make
the Raspbian image available as a block device in the filesystem. For this
command to work, the tested administrator distribution must have the
\texttt{util-linux} package with version 2.21 or higher. Otherwise, the
\texttt{-P} argument of \texttt{losetup} will appear as invalid. If the
version of \texttt{losetup} can not be updated for some reason, an
alternative option for this part is presented in
Section~\ref{sec:alternative_losetup} of the Appendix.

@@

The image block device appears as \texttt{/dev/loop0}. This block device has
two partitions associated to it, e.g. \texttt{loop0p1} and \texttt{loop0p2}.
Finally, we check the block device status with \texttt{e2fsck} and
\texttt{resize2fs}, respectively in the command block:

@@

If successfully executed, our terminal should have changed the prompt
indicating that we are the root user in the Raspbian lite root filesystem as
the apparent root. In case the \texttt{chroot} command is not successful,
we provide an alternative command in Section~\ref{sec:chroot} of the Appendix.
To be aware of the mode that we are working now, we change
the prompt title to indicate that it is a \texttt{chroot} environment as
follows:

>>>>

\label{sec:image_setup}
In this section, we review the steps to create a common \ac{OS} image for
all the \ac{Raspi}s. The image setup is composed of three major steps:
Select and download the \ac{OS} image file, alter the image structure and
configure the \ac{OS} files. We proceed to detail all these steps providing
brief discussions to our setup choices when required. To perform these
steps, we indicate with command-line blocks the required sequential
commands to be typed by the testbed administrator in his / her \ac{PC}
to obtain the desired setting. In all the following sections, the number
sign (\#) and dollar sign (\$) will be used in the command blocks in the
paper to indicate if an operation needs to be run with root permissions or
common user permissions, respectively. The signs will be posted as the
terminal prompt in the command blocks.

@@

In this code block, the \texttt{\$\{URL\}} and \texttt{\$\{IMAGE\}}
variables specify where the Linux bundle is located and
\texttt{\$\{WORKDIR\}} specifies a working directory where the Raspbian
Lite bundle will be downloaded and customized. Notice that even though we
use the \$ and \# signs in the shell, in general these signs will be
particular to the testbed administrator \ac{OS} shell. Next, we create
the working directory and change to it with the \texttt{cd} command. To
download the image, we utilize the \texttt{wget} command before unpacking
the \texttt{zip} file as follows:

@@

After Raspbian Lite has been unpacked, there should be an \texttt{.img}
file in the working directory \texttt{\$\{WORKDIR\}}. Here, \texttt{fdisk} can
be used to display the content of the image. We parse the arguments \texttt{-u sectors}
to display the sizes in sectors and \texttt{-l} to display the partitions
within the image. The \texttt{fdisk} command should output to the terminal
something similar to:

@@

\begin{lstlisting}[]
Command (m for help): Partition number (1-4):
Command (m for help): Partition type:
   p   primary (1 primary, 0 extended, 3 free)
   e   extended
Select (default p): Using default response p
Partition number (1-4, default 2): Using default value 2
First sector (2048-4806655, default 2048): Last sector, +sectors or
+size{K,M,G} (137216-4806655, default 4806655): Using default value 4806655

@@

\subsection{Loopback Device Setup}
After successfully resizing the image file, we use a loopback device to make
the Raspbian image available as a block device in the filesystem. For this
command to work, the tested administrator distribution must have the
\texttt{util-linux} package with version 2.21 or higher. Otherwise, the
\texttt{-P} argument of \texttt{losetup} will appear as invalid. If the
version of \texttt{losetup} can not be updated for some reason, an
alternative option for this part is presented in
Section~\ref{sec:alternative_losetup} of the Appendices.

@@

The image block device appears as \texttt{/dev/loop0}. This block device has
two partitions associated to it, e.g. \texttt{loop0p1} and \texttt{loop0p2}.
Finally, we check the filesystem of the block device with \texttt{e2fsck} and
resize it with the \texttt{resize2fs} command:%, respectively in the command block:

@@

If successfully executed, our terminal should have changed the prompt
indicating that we are the root user in the Raspbian lite root filesystem as
the apparent root. In case the \texttt{chroot} command is not successful,
we provide an alternative command in Section~\ref{sec:chroot} of the Appendices.
To be aware of the mode that we are working now, we change
the prompt title to indicate that it is a \texttt{chroot} environment as
follows:

### 2016-08-25 introduction.tex
more corrections

The upcoming 5G technology is targeting the controlling and steering of the Internet of Things (IoT) in real-time on global scale. This will break new ground for new markets such as driverless vehicles, manufacturing, humanoid robots, smart grids, and many more. With 5G, the number of wireless devices will increase by five times up to 50 billion devices \cite{cisco2011forecast}. It is generally acceptable, that those devices will not be connected in the same cellular manner as the current 10 billion mobile devices are connected due to the expected energy consumption. Therefore the communication architecture in future communication scenario will be dominated by mesh technologies. Mesh technology has been known for sensor and ad hoc networks or mobile cloud scenarios, but the technical requirements on 5G mesh-based communication systems are dramatically increasing. Future mesh networks need to support high data rate, low latency, security, network availability and heterogeneous devices. In state of the art systems those requirements are traded-off with each other, but in the 5G context we cannot do this anymore.

@@

The emergence of powerful and inexpensive single-board computers running full versions of a standard \ac{OS} from many possible opens new possibilities in this area. This not only allows to seamlessly develop implementations that are compatible with higher end devices and allows the reuse stable software from widely-used \ac{OS}. Given this set of specific needs, in this work we present the design, key step-by-step instructions and mechanisms to setup, configurate and maintain of an inexpensive testbed using Raspberry Pi devices for networking (wireless or wired) and storage applications including \ac{RLNC} functionalities into the testbed through Kodo. Nevertheless, the architecture itself is not bounded to the networking area and can be used for other applications that require replicable results with Raspberry Pi. Our work for the testbed procedure is organized as follows: Section~\ref{sec:overview} introduces the testbed system. In Section~\ref{sec:image_setup}, we provide details about the testbed setup, scripts, configuration files and connectivity. In Section~\ref{sec:overlay_fs}, we elaborate on the need and setup for an overlay filesystem for our testbed in order to have both persistent and non-persistent data on it as an optional step. Section~\ref{sec:tools} describes a set of automation and monitoring tools that can be included in the testbed to simplify the execution of routinary and repetitive tasks. Section~\ref{sec:cross_compilation} elaborates on the compilation of the Kodo library for the \ac{Raspi}. Conclusions and future work are reviewed in Section~\ref{sec:conclusions}. Finally, a set of alternative commands, in case the ones presented in this work might not be executed, are discussed in the Appendix.

>>>>

The upcoming 5G technology is targeting the controlling and steering of the Internet of Things (IoT) in real-time on global scale. This will break new ground for new markets such as driverless vehicles, manufacturing, humanoid robots, smart grids, and many more. With 5G, the number of wireless devices will increase by five times up to 50 billion devices \cite{cisco2011forecast}. It is generally believed, that those devices will not be connected in the same manner as current devices are connected today. Centralized system will collapse in terms of capacity, while distributed systems appear as an alternative. Therefore the communication architecture in future systems will be dominated by mesh technologies. Mesh technology has been known for sensor and ad hoc networks or mobile cloud scenarios, but the technical requirements on 5G mesh-based communication systems are dramatically increasing. Future mesh networks need to support high data rate, low latency, security, network availability and heterogeneous devices. In state of the art systems those requirements are traded-off with each other, but in the 5G context we cannot do this anymore.

@@

The emergence of powerful and inexpensive single-board computers running full versions of a standard \ac{OS} from many possible opens new possibilities in this area. This not only allows to seamlessly develop implementations that are compatible with higher end devices and allows the reuse stable software from widely-used \ac{OS}. Given this set of specific needs, in this work we present the design, key step-by-step instructions and mechanisms to setup, configurate and maintain of an inexpensive testbed using \ac{Raspi}~\cite{making_of_pi} devices for networking (wireless or wired) and storage applications including \ac{RLNC} functionalities into the testbed through Kodo. Nevertheless, the architecture itself is not bounded to the networking area and can be used for other applications that require replicable results with \ac{Raspi}~\cite{Cox2014,s141224408,s140509290,s16020220}\textcolor{red}{move citations or rewrite a bit here}. Our work for the testbed procedure is organized as follows: Section~\ref{sec:overview} introduces the testbed system. In Section~\ref{sec:image_setup}, we provide details about the testbed setup, scripts, configuration files and connectivity. In Section~\ref{sec:overlay_fs}, we elaborate on the need and setup for an overlay filesystem for our testbed in order to have both persistent and non-persistent data on it as an optional step. Section~\ref{sec:tools} describes a set of automation and monitoring tools that can be included in the testbed to simplify the execution of routinary and repetitive tasks. Section~\ref{sec:cross_compilation} elaborates on the compilation of the Kodo library for the \ac{Raspi}. Conclusions and future work are reviewed in Section~\ref{sec:conclusions}. Finally, a set of alternative commands, in case the ones presented in this work might not be executed, are discussed in the Appendices.

### 2016-08-25 overlay_fs.tex
more corrections

Assuming that we are still in the \texttt{chroot} environment into the
Raspbian Lite root filesystem for installing packages, we can setup the
overlay filesystem at this point of the procedure. There already exists
implementations overlaying the root filesystem. We use an implementation
available at the Git repository in \cite{overlayroot}. Since we have
installed \texttt{git} in a previous step, we clone the repository. The
command block below stores it in \texttt{/tmp} which is really mounted in
\ac{RAM}. Thus, all the files stored here will disappear when the system is
rebooted.

@@

For the \texttt{----recursive} option to work properly, it is necessary that
the package \texttt{util-linux} version is greater than or equal to 2.22.
Otherwise, an alternative is to either update the package or follow the
procedure in Section~\ref{sec:umount} of the Appendix.

@@

If the last command was executed correctly, the
\texttt{\${ROOTDIR}/etc/fstab} file should have the new line. However,
the line will not be properly indented. To verify the correct line
insertion and fix the indentation, edit the file with
\texttt{sudo vim \$\{ROOTDIR\}/etc/fstab}. The resulting file should like
the following:

@@

Again, if the \texttt{----recursive} option is not available, the
procedure in Section~\ref{sec:umount} of the Appendix could be followed.
If the steps are successfully executed up to this point, the customized image
is available in the \texttt{\$\{IMAGE\}.img} file and ready to be deployed
into the \ac{Raspi}s. In the following section, we indicate how to proceed
with the writing of the image into various memory cards.

\subsection{Writing Customized Image to SD Memory Cards}
For a basic system setup, the final step is to write the customized image
to all the memory cards before they can be used in the \ac{Raspi}s. For
our current considered system, we do this manually for each card.
The testbed administrator needs to insert each memory card in its \ac{PC}
and follow the procedure in this section. A given card will be available as
\texttt{/dev/mmcblkX} for memory card readers or as \texttt{/dev/sdbX}
USB-through readers, where \texttt{X} a natural number.

It is \textit{very important} to write to the correct device as
everything will be overwritten. To avoid removing information from the
wrong device, a testbed administrator can insert and eject a memory
card a couple of times and using the \texttt{sudo lsblk} or \texttt{df -h}
commands to deduce the correct device. For our case, the device was
\texttt{/dev/sdb} in an Ubuntu distribution and
\texttt{/dev/mmcblk0} in an Archlinux distribution. Once identified, to write
the image to a memory card, the following command is used:

@@

The previous \texttt{dd} and \texttt{sync} commands for copying the image to
the SD memory card and flushing the remainder in memory to the filesystem
will take around 15 minutes to be finished. This is an average on a two-core
1.5 GHz laptop. However, other times might occur depending on the total image
size, the writing block size and the core capabilities of the testbed
administrator \ac{PC}. After this is made, it is only required
to eject the memory card and now plug it in a \ac{Raspi} so it can boot up.

>>>>

Assuming that we are still in the \texttt{chroot} environment of the
Raspbian Lite root filesystem for installing packages, we can setup the
overlay filesystem at this point of the procedure. There already exists
implementations overlaying the root filesystem. We use an implementation
available at the Git repository in \cite{overlayroot}. Since we have
installed \texttt{git} in a previous step, we clone the repository. The
command block below stores it in \texttt{/tmp} which is really mounted in
\ac{RAM}. Thus, all the files stored here will disappear when the system is
rebooted.

@@

For the \texttt{----recursive} option to work properly, it is necessary that
the package \texttt{util-linux} version is greater than or equal to 2.22.
Otherwise, an alternative is to either update the package or follow the
procedure in Section~\ref{sec:umount} of the Appendices.

@@

If the last command was executed correctly, the
\texttt{\$\{ROOTDIR\}/etc/fstab} file should have the new line.
%However,
%the line will not be properly indented. To verify the correct line
%insertion and fix the indentation, edit the file with
%\texttt{sudo vim \$\{ROOTDIR\}/etc/fstab}.
The resulting file should look like the following:

@@

Again, if the \texttt{----recursive} option is not available, the
procedure in Section~\ref{sec:umount} of the Appendices could be followed.
If the steps are successfully executed up to this point, the customized image
is available in the \texttt{\$\{IMAGE\}.img} file and ready to be deployed
into the \ac{Raspi}s. In the following section, we indicate how to proceed
with the writing of the image into various memory cards.

\subsection{Writing Customized Image to SD Memory Cards}
For a basic system setup, the final step is to write the customized image
to all the memory cards before they can be used in the \ac{Raspi}s. For
our current considered system, we do this manually for each card.
The testbed administrator needs to insert each memory card in its \ac{PC}
and follow the procedure in this section. A given card will be available as
\texttt{/dev/mmcblkX} or \texttt{/dev/sdYX} where \texttt{X} is a natural number
and \texttt{Y} is a letter.

It is \textit{very important} to write to the correct device as
everything will be overwritten. To avoid removing information from the
wrong device, a testbed administrator can use the commands \texttt{lsblk}
and/or \texttt{df -h} before and after inserting the memory card to deduce
its correct device name.
For our case, the device was \texttt{/dev/mmcblk0}. Once identified, to write
the image to a memory card, the following command is used:

@@

The previous \texttt{dd} and \texttt{sync} commands for copying the image to
the memory card and flushing the remainder in memory to the filesystem
will take tens of minutes depending on the memory card speed and the size of
the image. After this is made, it is only required
to eject the memory card and now plug it in a \ac{Raspi} so it can boot up.

%The previous \texttt{dd} and \texttt{sync} commands for copying the image to
%the memory card and flushing the remainder in memory to the filesystem
%will take around 15 minutes to be finished. This is an average on a two-core
%1.5 GHz laptop. However, other times might occur depending on the total image
%size, the writing block size and the core capabilities of the testbed
%administrator \ac{PC}. After this is made, it is only required
%to eject the memory card and now plug it in a \ac{Raspi} so it can boot up.

### 2016-08-25 overview.tex
more corrections

A sketch of the testbed is depicted in Fig.~\ref{fig:testbed_setup}.
The testbed consists of up to 100 \ac{Raspi}s of different models.
More specifically, in our design we consider: Raspberry Pi 1 model B rev.~2,
Raspberry Pi 2 model B V1.1 and Raspberry Pi 3 model B V1.2.
All \ac{Raspi}s are each equipped with a 8 GB memory card, a wired and wireless
network interface and a power supply. All the \ac{Raspi} are connected to
a common \ac{LAN} that provides internal and external connectivity. Without
loss of generality, in our case they are connected to a university network
using their wired Ethernet interface that is named \texttt{eth0} according
to the legacy naming convention of Ethernet interfaces in
Linux~\cite{PredictableNetworkInterfaceNames}. We consider the
university network since our testbed is used by students and
employees to perform measurements and experimentation of controlled and
reproducible scenarios as part of academic research. Nevertheless, the
testbed description and procedures for setting it up are not restricted
to the academic scenario. All \ac{Raspi}s are configured to run
a \ac{SSH} daemon for easy remote access within the university network.
We requested the university \ac{IT} department to configure the university
\ac{DHCP} server to assign each \ac{Raspi} a static \ac{IP} address. This
eliminates the demand for monitors and keyboards with the \ac{Raspi}s
for non-graphical applications. Finally, our design aims to configure all
\ac{Raspi}s identically from a customized bootable image in their
respective memory cards, while still allowing the end-users to store
files locally in each of the \ac{Raspi}s.

@@

In Fig.~\ref{fig:testbed_setup} and what follows, we will refer as the
testbed administrator to the person(s) in charge of setting up and
configuring the testbed with administrator privileges from the \ac{OS}
point of view. The setting and configuration procedure commands are
performed by the testbed administrator in a regular \ac{PC} running a
Linux distribution. Although in principle the administrator
Linux distribution is not a restriction, we present our procedure in a
Debian-based Linux distribution. Our basic design considers to create
a customized image to later store it in a memory card for each \ac{Raspi}.
Once configured, we store the resulting image file in a \ac{HTTP} server
as backup and in case the testbed administrator requires to make new
changes to this file. We also put all the required configuration files
and scripts for the \ac{Raspi}s setup in the \ac{HTTP} server so there is
a single place where system setup is stored and could be modified.
This simplifies the system maintenance at a later stage.
As another design consideration, when working on a testbed
with many devices, it may not always be desired to make persistent changes
on them. One scenario is when different users are interested in
running experiments on a rebooted testbed. Thus, we later present how
to utilize stacked filesystems to enable both persistent and
temporary storage to have this capability. Its purpose is to
remove non-desired data after a reboot while keeping the original
customized image structure. This step of the procedure is optional if
the testbed administrator decides to keep only persistent changes regardless
of the testbed use. Finally, we include a set of automation, monitoring
and cross-compilation tools over the top of our system in order to simplify
the execution of repetitive and long tasks, be able to follow the progress
of long task processes and compile relevant C++ source code for the testbed
administrator.

>>>>

A sketch of the testbed is depicted in Fig.~\ref{fig:testbed_setup}.
The testbed consists of up to 100 \ac{Raspi}s of different models.
More specifically, in our design we consider: Raspberry Pi 1 model B rev.~2,
Raspberry Pi 2 model B V1.1 and Raspberry Pi 3 model B V1.2.
All \ac{Raspi}s are each equipped with a 8 GB \ac{SD} memory card, a wired and wireless
network interface and a power supply. All the \ac{Raspi} are connected to
a common \ac{LAN} that provides internal and external connectivity. Without
loss of generality, in our case they are connected to a university network
using their wired Ethernet interface that is named \texttt{eth0} according
to the legacy naming convention of Ethernet interfaces in
Linux~\cite{PredictableNetworkInterfaceNames}. We consider the
university network since our testbed is used by students and academic staff
to perform measurements and experimentation of controlled and
reproducible scenarios as part of academic research. Nevertheless, the
testbed description and procedures for setting it up are not restricted
to the academic scenario. All \ac{Raspi}s are configured to run
a \ac{SSH} daemon for easy remote access within the university network.
We requested the university \ac{IT} department to configure the university
\ac{DHCP} server to assign each \ac{Raspi} a static \ac{IP} address. This
eliminates the demand for monitors and keyboards with the \ac{Raspi}s
for non-graphical applications. Finally, our design aims to configure all
\ac{Raspi}s identically from a customized bootable image in their
respective memory cards, while still allowing the end-users to store
files locally in each of the \ac{Raspi}s.

@@

We will refer to the
testbed administrator as the person(s) in charge of setting up and
configuring the testbed with administrator privileges from the \ac{OS}
point of view. The setting and configuration procedures are
performed by the testbed administrator in a regular \ac{PC} running a
Linux distribution as shown in Fig.~\ref{fig:testbed_setup}. Although in principle the administrator
Linux distribution is not a restriction, we present our procedure in a
Debian-based Linux distribution. Our basic design considers to create
a customized image to store it later in a memory card for each \ac{Raspi}.
Once configured, we store the resulting image file in a \ac{HTTP} server
as backup and in case the testbed administrator requires to make new
changes to this file. We also put all the required configuration files
and scripts for the \ac{Raspi}s setup in the \ac{HTTP} server so there is
a single place where system setup is stored and could be modified.
This simplifies the system maintenance at a later stage.
As another design consideration, when working on a testbed
with many devices, it may not always be desired to make persistent changes
on them. One scenario is when different users are interested in
running experiments on a rebooted testbed. Thus, we later present how
to utilize stacked filesystems to enable both persistent and
temporary storage to have this capability. Its purpose is to
remove non-desired data after a reboot while keeping the original
customized image structure. This step of the procedure is optional if
the testbed administrator decides to keep only persistent changes regardless
of the testbed use. Finally, we include a set of automation, monitoring
and cross-compilation tools over the top of our system in order to simplify
the execution of repetitive and long tasks, be able to follow the progress
of long task processes and compile relevant C++ source code for the testbed
administrator.

### 2016-08-25 tools.tex
more corrections

\subsection{Fabric}
Controlling multiple devices using \ac{SSH} from a single \ac{PC}
often leads to many repetitive tasks. Among these, we can mention:
(i) rebooting a set of devices, (ii) installing applications in multiple
devices and (iii) copying files to and / or from multiple devices, etc. To
simplify the management and avoid execution issues, Fabric provides
a Python library to ease the major task of working with multiple devices from
a single \ac{PC}. First, the testbed administrator creates a directory to
hold the \texttt{Fabric} source code:

@@

env.hosts = ['rasp00.domain.com','rasp01.domain.com','rasp02.domain.com']
env.user = 'pi'
env.password = 'raspberry'

@@

The previous \texttt{fabfile} shows three functions that perform our
example tasks. These functions utilize variables and subsequent functions
from the Fabric \ac{API} such as \texttt{env}, \texttt{task} and \texttt{sudo}
among others. Each of these \ac{API} functions permit to define environment
variables, create the administrator tasks through decorators or run the
mentioned task in \texttt{sudo} mode, respectively.
When a task is called from the terminal, Fabric searches the directory for
the \texttt{fabfile.py} file and executes the desired task. The syntax for
executing a task with arguments is in the form
\texttt{fab <TASK>:arg1,arg2,...}. In what follows, we denote the
\ac{IP} address of a generic \ac{Raspi} for test as \texttt{<RASP\_IP>}.
The executions from the terminal of some of these commands are shown as
follows:

@@

The first function above reboots the \ac{Raspi}s in the lists of hosts and
the second function installs a program given by an argument. For the
connection to the devices, Fabric calls the Paramiko module from Python
to make a \ac{SSH} connection. For this to work properly, the Paramiko
version needs to be higher than or equal to 1.15.1. If not available,
the \ac{SSH} connections from Fabric may fail. Thus in case of any problems,
some instructions for updating the Paramiko package are available in
Section~\ref{sec:paramiko} of the Appendix. This is a standard recommendation
the Fabric troubleshooting guide~\cite{2016fabricsupport}.

@@

After attaching again, we should see \texttt{top} still running.
\texttt{screen} has more functionalities that can be used in this
or other contexts but that is outside the scope of this work.
To terminate the \texttt{screen} session, first terminate \texttt{top}
by pressing \texttt{q} in the keyboard.
Once \texttt{top} is terminated,
%Once out of the \texttt{top} output,
we need to type \texttt{exit} two times to first exit the
\texttt{screen} session and then terminate the \ac{SSH} connection.
An output should be as follows:

>>>>

\subsection{Fabric}
Controlling multiple devices using \ac{SSH} from a single \ac{PC}
often leads to many repetitive tasks. Among these, we can mention:
(i) rebooting a set of devices, (ii) installing applications in multiple
devices and (iii) copying files to and / or from multiple devices and so on. To
simplify the management and avoid execution issues, Fabric provides
a Python library to ease the major task of working with multiple devices from
a single \ac{PC}. First, the testbed administrator creates a directory to
hold the \texttt{Fabric} source code:

@@

env.hosts = ['10.0.0.100','10.0.0.101','10.0.0.102']
env.user = 'pi'
env.password = 'raspberry'

@@

The previous \texttt{fabfile} shows three functions that perform our
example tasks. These functions utilize variables and subsequent functions
from the Fabric \ac{API} such as \texttt{env}, \texttt{task} and \texttt{sudo}
among others. Each of these \ac{API} functions permits to define environment
variables, create the administrator tasks through decorators or run the
mentioned task in \texttt{sudo} mode, respectively.
When a task is called from the terminal, Fabric searches the directory for
the \texttt{fabfile.py} file and executes the desired task. The syntax for
executing a task with arguments is in the form
\texttt{fab <TASK>:arg1,arg2,...}. In what follows, we denote the
\ac{IP} address of a generic \ac{Raspi} for test as \texttt{<RASP\_IP>}.
The executions from the terminal of some of these commands are shown as
follows:

@@

The first function above reboots the \ac{Raspi}s in the lists of hosts and
the second function installs a program given by an argument. For the
connection to the devices, Fabric calls the Paramiko module from Python
to make a \ac{SSH} connection. For this to work properly, the Paramiko
version needs to be higher than or equal to 1.15.1. If not available,
the \ac{SSH} connections from Fabric may fail. Thus in case of any problems,
some instructions for updating the Paramiko package are available in
Section~\ref{sec:paramiko} of the Appendices. This is a standard recommendation
the Fabric troubleshooting guide~\cite{2016fabricsupport}.

@@

After attaching again, we should see \texttt{top} still running.
\texttt{screen} has more functionalities that can be used in this
or other contexts but that is outside the scope of this work.
To terminate the \texttt{screen} session, first terminate \texttt{top}
by pressing \texttt{q} in the keyboard.
Once \texttt{top} is terminated,
%Once out of the \texttt{top} output,
we need to type \texttt{exit} two times in order to first exit the
\texttt{screen} session and then terminate the \ac{SSH} connection.
An output should be as follows:
